---
title: "Practice using ENM functions in kuenm"
author: "Marlon E. Cobos"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    fig_width: 7
    fig_height: 6
    toc_depth: 3
    use_bookdown: true
    number_sections: false
---

```{r setup, include=FALSE}
# global options
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, prompt = FALSE, tidy = TRUE,
                      comment = NA, message = FALSE, warning = FALSE, 
                      fig.align = "center")
knitr::opts_knit$set(width = 100)
```

## Description

This script is a guide to. 

The data that we would use can be downloaded following this **link**. 

<br>

## Setting up R

### Initial dependencies



### R pacakages needed

The following lines of code help to install the R package needed to perform ENM and all relevant dependencies.  

```{r install, eval=FALSE}
# R package to install packages from GitHub
install.packages("remotes")

# R package to run dispersal simulations
remotes::install_github("marlonecobos/kuenm")
```

Now we can load the packages needed to read the data we will use and run the simulations.

```{r load, message=FALSE}
# loading packages
library(raster)
library(kuenm)
```


### Working directory 

The simulations will be run using data from a local directory, and results will also be written in the this directory. For these reasons, it is important to define a working directory that suits your needs.

```{r wd, eval=FALSE}
# please change "YOUR/DIRECTORY" as needed
setwd("YOUR/DIRECTORY")
```

<br>

## Understanding ENM workflow

### Data required

### Workflow description

The information required to run this type of simulation is related to  

<br>

## Example case

### Getting the data

