---
title: "Exploring variability and uncertainty in ENM results"
author: "Marlon E. Cobos"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    fig_width: 7
    fig_height: 6
    toc_depth: 3
    use_bookdown: true
    number_sections: false
---

```{r setup, include=FALSE}
# global options
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, prompt = FALSE, tidy = TRUE,
                      comment = NA, message = FALSE, warning = FALSE, 
                      fig.align = "center")
knitr::opts_knit$set(width = 100)
```

## Description

This is a guide to assess and map variability and uncertainty in results from ecological niche modeling (ENM) exercises. We use the <a href="https://github.com/marlonecobos/kuenm#kuenm-an-r-package-for-detailed-development-of-maxent-ecological-niche-models" target="_blank">kuenm</a> package to perform analyses using results from final model and projections. The guide is focused on the functions of the package dedicated to post-modeling analysis.  

The data and results required to perform the analyses in this guide can be obtained by following the **kuenm_practice guide**, which can be in the GitHub repository <a href="https://github.com/marlonecobos/ASM_course" target="_blank">ASM Course</a>.

<br>

## R pacakge and working directory

Now we can load the package needed and set the working directory.

```{r load_wd, eval=FALSE}
# loading packages
library(kuenm)

# set working directory, please change "YOUR/DIRECTORY" as needed
setwd("YOUR/DIRECTORY")
```

<br>

## Workflow description

The information required to run this type of simulation is related to  

<br>

## Example case

### Mapping variance from distinct sources

```{r var, eval=FALSE}
# function documentation and argument definition
help(kuenm_modvar)

spname <- "Amblyomma_americanum"
moddir <- "Final_models"
format <- "asc"
replicated <- TRUE
curpatt <- "current"
projected <- TRUE
emscepatt <- c("rcp45", "rcp85")
gcms <- c("cc", "mi")
ext_selected <- "E"
dir_var <- "Variance_maps
```

Now the analysis.

```{r var1, eval=FALSE}
# mapping the variance from distinct sources
kuenm_modvar(sp.name = spname, fmod.dir = moddir, is.swd = swd, 
             replicated = replicated, format = format, project = projected, 
             current = curpatt, emi.scenarios = emscepatt, 
             ext.type = ext_selected, clim.models = gcms, out.dir = dir_var)
```

### Hierarchical partitioning of variance

```{r hierp, eval=FALSE}
# function documentation and argument definition
help(kuenm_hierpart)

iterations <- 100
sample_size <- 1000
hpart_dir <- "Hierarchical_partitioning"
```

Let's run the analysis.

```{r hierp1, eval=FALSE}
# mop analysis for multiple scenarios
kuenm_hierpart(sp.name= spname, fmod.dir = moddir, replicated = replicated, 
               format = format, project = projected, current = curpatt,
               emi.scenarios = emscepatt, clim.models = gcms, 
               iterations = iterations, sample.size = sample_size,
               ext.type = ext_selected, factors.col, out.dir = hpart_dir)

```


### MOP analysis

The mobility-oriented parity metric (MOP; Owens et al. 2013) allows to identify levels of similarity between environmental conditions in the calibration area and other scenarios where models will be projected. The results obtained can be mapped to detect areas where conditions are completely different (non-analogous) to those in the calibration area (areas of strict extrapolation). Interpretations in areas of strict extrapolation must be done carefully as the results in those areas are the result of how models are extrapolating. The package kuenm contains tools to perform the MOP analysis for distinct scenarios. 

```{r mop, eval=FALSE}
# function documentation and argument definition
help(kuenm_mmop)

set <- "Set_4"
swd <- FALSE
percentage <- 10
comp <- 2000
parall <- TRUE
outmop <- "MOP_results"
```

Now the analysis.

```{r mop1, eval=FALSE}
# mop analysis for multiple scenarios
kuenm_mmop(G.var.dir = gvars, M.var.dir = mvars, is.swd = swd, sets.var = set, 
           out.mop = outmop, percent = percentage, comp.each = comp, 
           parallel = parall)
```

### MOP agreement

When multiple scenarios and GCMs are used, it may be a good idea to calculate the agreement of areas detected to be non-analogous among those GCMs for every scenario in question. The following lines of code will help to identify such areas of agreement.

```{r mopag, eval=FALSE}
# function documentation and argument definition
help(kuenm_mopagree)

format1 <- "GTiff"
mopag_dir <- "MOP_agreement"
```

Now the calculations.

```{r mopag1, eval=FALSE}
# agreement of areas of extrapolation
kuenm_mopagree(mop.dir = outmop, in.format = format1, out.format = format1, 
               current = curpatt, emi.scenarios = emscepatt, clim.models = gcms,
               out.dir = mopag_dir)
```
